---
title: "Final project2"
author: "Yuchen li"
date: "2025-02-21"
output: html_document
---
# Introduction
# This analysis explores the key factors influencing diabetes using the Pima Indians Diabetes Dataset.  
# We examine correlations between glucose, BMI, blood pressure, insulin, age, and diabetes outcomes  
# to identify significant risk factors and visualize their relationships.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(mice)
library(car)
library(caret)
library(ggplot2)
library(knitr)
library(tidyr)
```

## Data description 
```{r}
file_path <- "data/diabetes.csv"
diabetes  <- read.csv(file_path)

# Data pre-processing

diabetes$BMI[diabetes$BMI == 0] <- NA
diabetes$Glucose[diabetes$Glucose == 0] <- NA
diabetes$BloodPressure[diabetes$BloodPressure == 0] <- NA

clean_diabetes <- diabetes %>% 
  filter(!is.na(Glucose)) %>% 
  filter(!is.na(BloodPressure)) %>% 
  filter(!is.na(BMI))

str(diabetes)
dim(diabetes)
summary(diabetes)
```

```{r}
summary_table <- clean_diabetes %>%
  summarise(
    Mean_Glucose = mean(Glucose, na.rm = TRUE),
    SD_Glucose = sd(Glucose, na.rm = TRUE),
    Mean_BMI = mean(BMI, na.rm = TRUE),
    SD_BMI = sd(BMI, na.rm = TRUE),
    Mean_BloodPressure = mean(BloodPressure, na.rm = TRUE),
    SD_BloodPressure = sd(BloodPressure, na.rm = TRUE),
    Mean_Age = mean(Age, na.rm = TRUE),
    SD_Age = sd(Age, na.rm = TRUE)
  )
summary_table
kable(summary_table, caption = "Summary of Key Variables")
```

### Summary Table Description

"The summary table presents the mean and standard deviation (SD) of key variables in the cleaned diabetes dataset. 
It provides an overview of the dataset's distribution and helps in understanding variable characteristics."
- **Glucose:** Mean = 121.88, SD = 30.75  
- **BMI:** Mean = 32.47, SD = 6.89  
- **Blood Pressure:** Mean = 72.40, SD = 12.38  
- **Age:** Mean = 33.35, SD = 11.76 


```{r}
correlation_table <- clean_diabetes %>%
  summarise(
    Cor_Glucose = cor(Glucose, Outcome, use = "complete.obs"),
    Cor_BMI = cor(BMI, Outcome, use = "complete.obs"),
    Cor_BloodPressure = cor(BloodPressure, Outcome, use = "complete.obs"),
    Cor_Insulin = cor(Insulin, Outcome, use = "complete.obs"),
    Cor_Age = cor(Age, Outcome, use = "complete.obs")
  )
correlation_table
kable(correlation_table, 
      caption = "Correlation between Key Variables and Diabetes Outcome",
      digits = 3)
```

### Correlation Table Description

"The correlation table shows the relationship between key variables and diabetes outcome.
Glucose has the strongest correlation with diabetes, followed by BMI, while other variables show weaker associations."
- **Glucose (0.488):** Moderate positive correlation with diabetes.  
- **BMI (0.299):** Weak positive correlation.  
- **Blood Pressure (0.166), Insulin (0.145), Age (0.246):** Weak correlations.

```{r}
long_data <- clean_diabetes %>%
  select(Glucose, BMI, BloodPressure, Insulin, SkinThickness, Age, Outcome) %>%
  pivot_longer(cols = -Outcome, names_to = "Variable", values_to = "Value")

ggplot(long_data, aes(x = as.factor(Outcome), y = Value, fill = as.factor(Outcome))) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free") + 
  theme_minimal() +
  labs(title = "Distribution of Key Variables by Diabetes Outcome",
       x = "Diabetes Outcome (0: No, 1: Yes)", y = "Value", fill = "Outcome")
```
### Boxplot Description

"The boxplot visualizes the distribution of key variables across diabetes outcomes (0: No diabetes, 1: Diabetes). 
It shows that individuals with diabetes tend to have higher glucose, BMI, and age compared to those without diabetes.
Blood pressure, insulin, and skin thickness also exhibit variations between the two groups, but with less pronounced differences."




